---
title: "Line Intercept Riparian Survey 3 <br>Riparian Cover"
author: "Oliver Franklin"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_download: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(fig.width=12, fig.height=9, echo = TRUE)
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")}
pacman::p_load(DT, tidyverse, abdiv, htmlwidgets, viridis,update=F)
```


```{r export functions and folder, echo = FALSE}
# functions to use for exporting results to output_folder (which is prompted in R when not in R environment)

#export figure (pdf for static ggplot plots, html for interactive plotly plots)
export_plot <- function(plot, filename) {
  if (inherits(plot, "ggplot")) {
    # Export ggplot as PDF
    ggsave(
      filename = file.path(output_folder, filename),
      plot = plot,
      device = "pdf",
      width = 11,
      height = 8.5
    )
  } else if (inherits(plot, "plotly")) {
    # Ensure the filename has .html extension
    html_filename <- file.path(output_folder, sub("\\.pdf$", ".html", filename))
    # Export plotly as a **single self-contained** HTML file
    saveWidget(
      widget = plot, 
      file = html_filename, 
      selfcontained = TRUE  # 
    )
  } else {
    stop("Unsupported plot type. Only ggplot and plotly objects are supported.")
  }
}

# Function to save a table
export_table <- function(table, filename) {
  write.csv(
    table,
    file = file.path(output_folder, filename),
    row.names = FALSE
  )
}

# Function to save summary as txt
export_summary <- function(summary_text, filename_base) {
  # Ensure summary_text is a character vector (it should already be, but just in case)
  summary_text <- as.character(summary_text)
    # Save as .txt
  txt_path <- file.path(output_folder, paste0(filename_base, ".txt"))
  writeLines(summary_text, con = txt_path)
}

# Examples of exporting individual results:
# Export a figure
# export_plot(ggplot_plot, "Thalweg Water Depth by Distance Upstream.pdf")
# export_plot(plotly_plot, "Thalweg Water Depth by Distance Upstream.html")
# 
# # Exporting a table
# export_table(as.data.frame(summary_table), "summary_table.csv")
# 
# Example of exporting a summary
# summary_text <- capture.output(summary(mtcars))
# export_summary(summary_text, "summary")

# below code to confirm an output folder is already specified, OR will request it is specified in R (interactively via readline)
# authors output folder: "~/Git/Core Monitoring/standardised protocols/data_tidier/Longitudinal"
# BUT if you want to knit, you will need to specify a default location (within the 'else' term)
if (interactive()) {
  while (!exists("output_folder") || !dir.exists(output_folder)) {
    output_folder <- readline(prompt = "Specify the output folder: ")
    if (!dir.exists(output_folder)) {
      tryCatch({
        dir.create(output_folder, recursive = TRUE)
        message("Created folder: ", output_folder)
      }, error = function(e) {
        message("Invalid folder. Please try again.")
        output_folder <- NULL
      })
    }
  }
} else {
  # Specify a default folder for non-interactive mode (e.g., knitting)
  output_folder <- "C:/Users/franklino/Documents/Restoration_Monitoring_Protocols/data_tidier/Nathan Creek/Riparian"
  if (!dir.exists(output_folder)) {
    dir.create(output_folder, recursive = TRUE)
    message("Default output folder created: ", output_folder)
  }
}
```

# Data import

For this script we import the tidied dataframe generated by the Data Wrangling script, which was exported to your output folder.

```{r import data, echo=FALSE}
importeddata <- file.path(output_folder, "LIRS_dataframe.csv")
df <- read.csv(importeddata)  

## if you need to specify a different location to import from, use below code:
# importeddata<-read.csv("~/Git/Core Monitoring/standardised protocols/data_tidier/Longitudinal/Long_dataframe.csv") # specify your file location & name here
# df<-data.frame(importeddata)
```

Here we will examine how densely vegetated the surveyed area is. We will explore the following:   

* Cover density across the surveyed riparian area;  
* Cover density across the riparian area, subset by height class (canopy/mid/low); and  
* Cover density of individual species within the riparian area.  

## Plant Species and 'Non-plant' Occurrences

Remember that our data includes both plant and 'non-plant' occurrences observed in the field. Consider where non-plant data needs to be excluded prior to summing or taking means of the data: For relative plant abundance, we recommended removing the non-plant occurrences. When exploring cover density, we typically remove the non-plant occurrences (e.g., to avoid including extents of riprap etc. in the total cover per m figure). 

However, there are cases where it may be of interest to explore the cover density of particular non-plant features. You may be interested in the extent of saturated soils in the riparian area. Or, perhaps for your study you want to exclude exposed bedrock to understand how vegetation changes in gravel deposits. The code chunk below can be modified to include non-plant occurrences, but by default we will exclude them and focus on density metrics based only on the plant species occurrences.

```{r include non-plants?}
include_non_plants <- FALSE   # change to TRUE to keep the non-plant occurrences (Plant_codes starting with "!")
```

## Overhanging Vegetation

Another consideration is whether or not to include overhanging vegetation in the cover metrics. Although vegetation that extends horizontally above the water surface is technically outside of the transect, we recommended that it was measured because of its functional importance to aquatic life. Consider whether you would want two otherwise identical transects to return different cover metrics if only one transect featured overhanging vegetation. By default, our code excludes overhanging vegetation from cover metrics, but extents of overhanging vegetation can be included using the code below (note that excluding overhanging vegetation does not exclude an entire plant, only the extent that is above the water surface / beyond the transect 0 m mark).

```{r include overhanging veg?}
include_overhanging <- FALSE   # change to TRUE to keep overhanging extents of vegetation (negative Start_location values)
```

```{r cover.df with user preferences, echo=FALSE}
cover.df <- df %>%
  { if (!include_non_plants) filter(., !str_starts(plant_code, "!")) else . } %>%
  {
    if (include_overhanging) {
      # keep all rows, even with negative End_location. Will convert to positive values if overhanging veg comes from beside transect, no value on land (Start plus End both negative values)
      mutate(., occurrence.magnitude = abs(End_location - Start_location))
    } else {
      # keep rows but adjust negative End_location and Start_location to 0 for calculating occurrence.mag (lots of zeros but keeps plants in data)
      mutate(.,occurrence.magnitude = pmax(End_location, 0) - pmax(Start_location, 0))
    }
  }
```

## Riparian Vegetation Cover Density

The density of vegetation cover is of interest because it provides information on the amount of vegetation present, rather than the proportions of a given species relative to other types of vegetation (relative abundance). Cover may be of particular importance in restoration because one would expect the amount of vegetation present to be associated with the riparian area's resilience against invasive plants, the contribution of leaf litter and insect drop to aquatic ecosystems, the filtering of pollutants, shading, etc. 

It is likely that your restoration objectives involve both achieving a certain level of plant density as well as a certain level of diversity derived from relative abundance data. So once you have understood the metrics and tests we generate in this script, you should consider any patterns that stand out alongside any changes in relative abundance from the preceding script. These types of interpretations are rarely simple, so we recommend getting together with your restoration partners to discuss.

Considering both cover density and plant community composition together helps to overcome a limitation of our data collection method. In short, you must be cautious when interpreting changes in cover that are also associated with significant shifts in the plant community composition. In order to avoid overly time consuming field surveys, we grouped together as a single occurrence any observations of stands of the same plant species that were contiguous and in the same height class (e.g., a monoculture of sedges might be many individual plants but would be recorded as one occurrence of sedge spanning _x_ metres). But if later you observed many other plants intermixed with the sedge, then we would expect to have recorded more overlaps of individual plants (e.g., each skunk cabbage established within a sedge stand overlaps to some degree). This additional overlap will increase the cover metrics. Whether or not this is of concern for your interpretation may depend on your focal cover objective.

### Total Cover Density (combined height classes)

In our data wrangling script we have already defined the length of your riparian transects, and we will use the extents of each plant species as a proportion of the transect length to generate cover metrics. First we will calculate the total cover density per transect, which incorporates all vegetation at any height class and probably best reflects what you think of as vegetation density. Note that the total cover (m vegetation per m riparian transect) is expected to be >1.0, due to the presence of overlapping multi-canopy vegetation (and overhanging vegetation, if included).

We will display the mean and sd values for each site and year below, and will export this table and the table with the values per transect to your output folder.

```{r cover per transect, echo=FALSE}
total.cover <- cover.df %>%
  group_by(Site, year, Transect_ID) %>%
  mutate(cover.per.m = occurrence.magnitude / Transect_length) %>%  # cover in m/m
  summarize(sum_cover_per_m = sum(cover.per.m, na.rm = TRUE), .groups = "drop")

# Detect whether Left/Right banks treated separately
has_banks <- any(grepl("Left|L|Right|R", total.cover$Transect_ID, ignore.case = TRUE))

if (has_banks) {
  total.cover <- total.cover %>%
    mutate(
      BankGroup = case_when(
        grepl("(Left|L)", Transect_ID, ignore.case = TRUE) ~ "Left Bank",
        grepl("(Right|R)", Transect_ID, ignore.case = TRUE) ~ "Right Bank",
        TRUE ~ "Combined"
      )
    )
} else {
  total.cover <- total.cover %>%
    mutate(BankGroup = "All")
}

# table for display - summary stats

summary_table <- total.cover %>%
  group_by(Site, year, BankGroup) %>%
  summarize(
    mean_val = mean(sum_cover_per_m, na.rm = TRUE),
    sd_val   = sd(sum_cover_per_m, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Summary = paste0(round(mean_val, 3), " (", round(sd_val, 3),") ")
  ) %>%
  select(Site, year, BankGroup, Summary)

datatable(
  summary_table,
  caption = htmltools::tags$caption(
  style = "caption-side: top; text-align: left; font-weight: bold;",
  "Mean and SD of plant cover density per linear m"),
  rownames = FALSE,
  options = list(
    columnDefs = list(list(targets = 3, className = "dt-right"))
  )
)

export_table(as.data.frame(summary_table),
             "Total plant cover density Summary Stats.csv")


# table for export only - transect values 

raw_table <- total.cover %>%
  mutate(sum_cover_per_m = round(sum_cover_per_m, 3)) %>%
  select(Site, year, BankGroup, Transect_ID, sum_cover_per_m)

export_table(as.data.frame(raw_table),
             "Total plant cover density Transect Data.csv")
```

It may also be interest to examine the density of vegetative cover at different height classes. Note that these different height classes are not independent - canopy density may increase as plant species composition and size changes, and higher densities in the canopy and midlayer influence the density and species composition below.

We will generate a figure per site-year showing density by height class and transect, which will be displayed below and exported to your output folder.

```{r cover by height, echo = FALSE, warning = FALSE}
cover.by.height <- cover.df %>%
  group_by(Site, year, Transect_ID, Height_class) %>%
  mutate(cover.per.m = occurrence.magnitude / Transect_length) %>%
  summarise(sum_cover_per_m = sum(cover.per.m, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    Height_class = dplyr::recode(Height_class,
                                 "L" = "Low",
                                 "M" = "Mid",
                                 "C" = "Canopy",
                                 .default = Height_class)   # keep "combo" unchanged
  )

desired_levels <- c("Low", "Mid","Canopy",  "combo") #orders them for display

# Ensure every Transect_ID × Height_class exists
cover.by.height <- cover.by.height %>%
  filter(!is.na(Height_class)) %>%
  complete(
    Site, year, Transect_ID,
    Height_class = intersect(desired_levels, unique(Height_class)),
    fill = list(sum_cover_per_m = 0)
  ) %>%
  mutate(
    Transect_ID = as.character(Transect_ID),
    Height_class = factor(Height_class, levels = desired_levels)
  )

# detect if banks treated separately
if (any(grepl("Left|L|Right|R", cover.by.height$Transect_ID, ignore.case = TRUE))) {
  cover.by.height <- cover.by.height %>%
    mutate(
      BankGroup = case_when(
        grepl("(Left|L)", Transect_ID, ignore.case = TRUE) ~ "Left Bank",
        grepl("(Right|R)", Transect_ID, ignore.case = TRUE) ~ "Right Bank",
        TRUE ~ "Combined"
      )
    )

  # Keep only banks that actually exist
  present_banks <- unique(cover.by.height$BankGroup)
  cover.by.height <- cover.by.height %>%
    filter(BankGroup %in% intersect(present_banks, c("Left Bank", "Right Bank")))
  has_banks <- TRUE
} else {
  cover.by.height$BankGroup <- "All Transects"
  has_banks <- FALSE
}

# display labels and numeric order (for y axis labels, that I could not get to work)
cover.by.height <- cover.by.height %>%
  group_by(Site,year) %>%
  mutate(
    Transect_num = as.numeric(str_extract(Transect_ID, "\\d+")), # grabs just numeric part (problematic if >1 numbers, e.g. "T12_1_R")
    Transect_label = paste0("T", Transect_num)
  ) %>%
  # Make Transect_ID a factor in numeric order
  arrange(Transect_num) %>%
  mutate(Transect_ID = factor(Transect_ID, levels = unique(Transect_ID)))

cover.by.height <- cover.by.height %>%
  mutate(label_x = -0.02 * max(sum_cover_per_m))%>%
  filter(!is.na(Height_class))

# Base plot
global_max<- max(cover.by.height$sum_cover_per_m, na.rm=TRUE) #used to make cover density axis same across plots

base_plot <- ggplot(cover.by.height,
                    aes(x = Height_class, y = sum_cover_per_m, fill = Transect_ID)) +
  geom_col(position = position_dodge(preserve = "single", width = 0.9)) +
  geom_text(aes(x = Height_class, y = label_x, label = Transect_label),
            position = position_dodge(width = 0.9),
            hjust = 1,
            size = 2,
            color = "black") +
  coord_flip() +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Vegetative cover density (m cover / m of transect) by height class",
    x = "Height Class",
    y = "Sum of cover per m"
  ) +
  scale_y_continuous(limits = c(-1, global_max)) +   # fixed y-axis is cover density axis (I flip coords for display)
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# split facets iff banks separated
if (has_banks) {
  coverbyheightplot <- base_plot +
    facet_wrap(~BankGroup, ncol = 2, scales = "fixed")
} else {
  coverbyheightplot <- base_plot
}

# Generate one plot per Site-year

# Ensure Site and year exist in cover.by.height
if (!("Site" %in% names(cover.by.height)) | !("year" %in% names(cover.by.height))) {
  stop("cover.by.height must contain Site and year columns.")
}

site_year_list <- unique(cover.by.height %>% select(Site, year))

plot_list <- list()

for (i in seq_len(nrow(site_year_list))) {

  this_site  <- site_year_list$Site[i]
  this_year  <- site_year_list$year[i]

  df_sy <- cover.by.height %>%
    filter(Site == this_site, year == this_year)

  # Base plot rebuilt for this subset
  p <- ggplot(df_sy,
              aes(x = Height_class, y = sum_cover_per_m, fill = Transect_ID)) +
    geom_col(position = position_dodge(preserve = "single", width = 0.9)) +
    geom_text(aes(x = Height_class, y = label_x, label = Transect_label),
              position = position_dodge(width = 0.9),
              hjust = 1, size = 2, color = "black") +
    coord_flip() +
    scale_fill_viridis_d(option = "D") +
    labs(
      title = paste0("Vegetative Cover Density (m/m)\n", this_site, " – ", this_year),
      x = "Height Class",
      y = "Sum of cover per m"
    ) +
    scale_y_continuous(limits = c(-1, global_max)) + 
    theme_minimal(base_size = 10) +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, face = "bold")
    )

  # If banks are separated, facet per bank
  if (has_banks) {
    p <- p + facet_wrap(~BankGroup, ncol = 2, scales = "fixed")
  }

  # Save plot to list
  plot_list[[paste0(this_site, "_", this_year)]] <- p

  # Export automatically
  export_plot(
    p,
    paste0("Veg Cover Density – ", this_site, " – ", this_year, ".pdf")
  )
}

plot_list 
```

And next, we will display the associated means and sd values for the height class plant density. The values per transect are exported to your output folder.

```{r table of cover by height, echo = FALSE}
# Detect which height classes are present in the data
present_classes <- intersect(desired_levels, unique(cover.by.height$Height_class))

# all transect table
cover_table_all <- cover.by.height %>%
  filter(!is.na(Height_class)) %>%
  ungroup() %>%
  select(-any_of(c("Transect_label", "label_x"))) %>%
  complete(Site, year, Transect_ID, Height_class = present_classes,
           fill = list(sum_cover_per_m = 0)) %>%
  pivot_wider(names_from = Height_class, values_from = sum_cover_per_m, values_fill = 0)

# Detect banks
has_banks <- any(grepl("Left|L|Right|R", cover_table_all$Transect_ID, ignore.case = TRUE))

if (has_banks) {
  cover_table_all <- cover_table_all %>%
    mutate(
      BankGroup = case_when(
        grepl("(Left|L)", Transect_ID, ignore.case = TRUE) ~ "Left Bank",
        grepl("(Right|R)", Transect_ID, ignore.case = TRUE) ~ "Right Bank",
        TRUE ~ "Combined"
      )
    )
}


# Generate Mean SD summary table (for display and export)
if (has_banks) {
  cover_summary_display <- cover_table_all %>%
    group_by(Site, year, BankGroup) %>%
    summarise(
      across(all_of(present_classes),
             ~ paste0(round(mean(.x, na.rm = TRUE), 3),
                      " (", round(sd(.x, na.rm = TRUE), 3), ") "),
             .names = "{.col}"),
      .groups = "drop"
    ) %>%
    rename(Transect_ID = BankGroup)
} else {
  cover_summary_display <- cover_table_all %>%
    group_by(Site, year) %>%
    summarise(
      Transect_ID = "Mean and SD",
      across(all_of(present_classes),
             ~ paste0(round(mean(.x, na.rm = TRUE), 3),
                      " (", round(sd(.x, na.rm = TRUE), 3), ") "),
             .names = "{.col}"),
      .groups = "drop"
    )
}

# Display Mean SD table
datatable(
  cover_summary_display,
  escape = FALSE,
  caption = "Vegetative cover density (Reach Mean and SD)",
  rownames = FALSE,
  options = list(
    columnDefs = list(list(targets = 1, className = "dt-right"))
  )
)

# --- Export both tables ---
export_table(as.data.frame(cover_summary_display), "Plant cover density MeanSD.csv")
export_table(as.data.frame(cover_table_all), "Plant cover density all transects.csv")
```


## Individual Species or Feature Cover Densities

We can also explore the cover density of individual species, or of the non-plant occurrences that you recorded. It may be of interest to (e.g.) compare the average extents of saturated soils, the amount of scoured bare substrates, or the density of a particular species such as an indicator species, a culturally important plant, or an invasive species of concern.

We will default to including the non-plant values while still excluding the overhanging vegetation, but this can be changed if you alter the code below, as before:

```{r include nonplants and overhangs?}
include_non_plants <- TRUE   # change to FALSE to remove the non-plant occurrences (Plant_codes starting with "!")
include_overhanging <- FALSE   # change to TRUE to keep overhanging extents of vegetation (negative Start_location values)
```

```{r cover.df with updated user preferences, echo=FALSE}
cover.df <- df %>%
  { if (!include_non_plants) filter(., !str_starts(plant_code, "!")) else . } %>%
  {
    if (include_overhanging) {
      # keep all rows, even with negative End_location. Will convert to positive values if overhanging veg comes from beside transect, no value on land (Start plus End both negative values)
      mutate(., occurrence.magnitude = abs(End_location - Start_location))
    } else {
      # keep rows but adjust negative End_location and Start_location to 0 for calculating occurrence.mag (lots of zeros but keeps plants in data)
      mutate(.,occurrence.magnitude = pmax(End_location, 0) - pmax(Start_location, 0))
    }
  }
```

```{r cover density by species, echo=FALSE, include = FALSE}
# Calculate per-transect plant cover density (m/m)
species.df <- cover.df %>%
  mutate(cover.per.m = occurrence.magnitude / Transect_length) 

# Aggregate cover by species, transect, and height class
plant_prop_transect <- species.df %>%
  group_by(Site, year, plant_code, Transect_ID, Height_class) %>%
  summarise(cover.per.m = sum(cover.per.m, na.rm = TRUE), .groups = "drop")

# Create all combinations of plant × transect × height (so later we can calculate means across all Transects)
all_combinations <- plant_prop_transect %>%
  distinct(Site, year, plant_code, Height_class) %>%
  inner_join(species.df %>% distinct(Site, year, Transect_ID),
             by = c("Site","year"), relationship="many-to-many") %>%
  tidyr::complete(Site, year, plant_code, Height_class, Transect_ID)

# Fill zeros where species-height not observed in a transect
plant_prop_transect_full <- all_combinations %>%
  left_join(plant_prop_transect,
            by = c("Site","year","plant_code","Height_class","Transect_ID")) %>%
  mutate(cover.per.m = replace_na(cover.per.m, 0))

# Now calculate mean and sd per site-year/species/height
species_height_summary <- plant_prop_transect_full %>%
  group_by(Site, year, plant_code, Height_class) %>%
  summarise(
    mean_cover = mean(cover.per.m, na.rm = TRUE),
    sd_cover   = sd(cover.per.m, na.rm = TRUE),
    .groups = "drop")

# Display table (non-zero rows only if desired)
display_table <- plant_prop_transect_full %>%
  filter(cover.per.m != 0)
  
#  Interactive datatable (showing nonzero values only)
datatable(
  display_table,
  options = list(
    columnDefs = list(
      list(
        targets = which(names(display_table) == "cover.per.m") - 1,
        render = JS("function(data, type, row, meta) {return parseFloat(data).toFixed(3);}")
      )
    )
  ),
  caption = "Cover density (m/m) of plant species by height class, measured per transect",
  rownames = FALSE
)


#Export full dataset (includes zeros for absent species)
export_table(as.data.frame(plant_prop_transect_full), "Plant cover density by species.csv")
```

The mean and sd density values for species/features by height cover across transects in the reach are displayed below. We have exported to your output folder this table as well as a table with the density values per species per transect.

```{r means sd species cover, echo=FALSE}
# check if banks treated separately
plant_prop_transect_full_bank <- plant_prop_transect_full %>%
  mutate(
    Bank = case_when(
      str_detect(Transect_ID, "Left|L") ~ "Left",
      str_detect(Transect_ID, "Right|R") ~ "Right",
      TRUE ~ NA_character_
    )
  )

bank_logic <- plant_prop_transect_full_bank %>%
  group_by(Site, year) %>%
  summarise(
    has_bank = any(!is.na(Bank)),
    .groups = "drop"
  )

plant_prop_transect_full_bank <- plant_prop_transect_full_bank %>%
  left_join(bank_logic, by = c("Site", "year")) %>%
  mutate(
    Bank_group = ifelse(has_bank, Bank, "Combined")  # If no Left/Right present → group = "Combined"
  )
# Add "Other" category for nonplants (entries without L/M/C height class)
plant_prop_transect_full_bank <- plant_prop_transect_full_bank %>%
  mutate(
Height_class_clean = case_when(
  Height_class %in% c("L", "M", "C") ~ Height_class,
  Height_class == "combo" ~ "combo",
  TRUE ~ "Other"
),
Height_class_clean = factor(Height_class_clean, levels = c("L","M","C","combo","Other"))
)

# Pivot wider to include height classes as columns
plant_prop_wide <- plant_prop_transect_full_bank %>%
  select(Site, year, Bank_group, plant_code, Transect_ID,
         Height_class_clean, cover.per.m) %>%
  pivot_wider(
    names_from = Height_class_clean,
    values_from = cover.per.m,
    values_fill = 0
  )

# Summary stats mean + SD
summary_stats <- plant_prop_wide %>%
  group_by(Site, year, Bank_group, plant_code) %>%
  summarise(
    across(
      any_of(c("L","M","C","combo","Other")),
      list(mean = ~mean(.x, na.rm = TRUE),
           sd   = ~sd(.x, na.rm = TRUE)),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

# Display table
datatable(
  summary_stats,
  options = list(
    pageLength = 25,
    columnDefs = list(
      list(
        targets = grep("mean|sd", names(summary_stats)) - 1,
        render = JS("function(data, type, row, meta) {return parseFloat(data).toFixed(3);}")
      )
    )
  ),
  caption = "Summary (mean and SD) of plant cover density by species and heightclass",
  rownames = FALSE
)

#
export_table(as.data.frame(summary_stats), "Plant cover density by species mean sd.csv")
```

## Inferential Tests

If you have a control or reference site and multiple years of data, we can apply inferential statistical tests to see if the observed changes are likely to be due to chance variability, or are consistent with the expected effects of your restoration actions. We will test whether there are statistically significant changes in the overall plant cover density values, using data from each transect grouped by Site and year, and we will also provide a method to test for significant changes in cover density for a given plant species (or 'non-plant' feature).

You will be familiar with p values of 0.05 (observed result expected just due to chance 1 in 20 times) and we will use p=0.05 as our cut-off here. However, depending on your adaptive management plans, you may be willing to accept different levels of certainty, and p values of 0.1 are not unheard of in restoration projects. This all depends on the risk tolerance of you and your team, and should be established before conducting analyses.

It is not appropriate to use ANOVA for this data because the cover density values are bounded by 0, are likely skewed, and because the data contain many zeroes (when a plant is present in certain transects, but not others, those other transects are assigned a zero value for that plant). Below we will use generalised linear models (GLMs), which are similar to linear regression but involve a link function and a non-normally distributed response variable. 

_NB_ We provide these tests as a general framework, and caution against using them if you do not understand what the models are doing. We may not have an opportunity to refine them (changes in government priorities) but consider them a useful placeholder for now. We will apply GLMs with a Gamma distribution (adjusting zeroes to 0.001) and a log link, because our response variable (cover density) should be positive, potentially zero-inflated, often skewed, and without a known upper bound (overlapping cover and height classes means densities exceeding 1m/m). However, actual data collected across a variety of ecosystems and stages of recovery might be very different, so we caution against taking the results of these GLMs at face value. If in doubt, our best advice is to rely on your own (and your restoration partners') interpretation of the raw data and the figures we have already presented above.

The tests are carried out below, after we specify the before and after data that you wish to compare.

#### Before / After Years

Enter your before and after years in the below code, you may have more than one year in each category. Remember that these should balance (equal number of years before as after), but they need not be consecutive years. If you do not have balanced data (e.g., only one year was possible before project implementation) be very cautious in running and interpreting the tests. You could select only a subset of your data to keep the comparison balanced, but if you do this for more than one set of dates you must account for the multiple comparisons inflating the likelihood of a Type I error (false positive).

```{r define before after}
# Define "Before" and "After" years
before_years <- c(2024) #add more years inside brackets if needed, e.g. 'c(2024, 2025, 2026)'
after_years <- c(2026) #add more years inside brackets if needed, e.g. 'c(2027, 2029, 2036)'
```

### Cover density: Generalised Linear Model (GLM)

The GLM will assess the statistical significance of observed changes in cover density, which is assessed at the level of each individual transect. The GLMs will be run for each height class separately, and for left and right banks separately if this has previously been specified. The interaction of Site*Time is interpretable as a change at your restoration site being distinct from a change at your control site. It is important to refer to your _a priori_ hypotheses and predictions before interpreting these results, as we will generate a number of tests and you must avoid interpreting them selectively (there might be significant results that appear by chance due to the number of tests being run). 

Because of the potential number of model outputs and diagnostic figures generated, we will suppress them from being displayed. Instead, we generate a summary table that only displays the Site*Time interaction model results, and we will export the full model results to your output folder. We do this because many people using this code will not want to interpret these models. If you do wish to interpret them, you should review the full models and the diagnostics (instructions within code).

```{r overall density glm, echo=FALSE, include=FALSE}
# to display the diagnostics in RStudio, you can change the preceding line where it says include=FALSE)


# update height-class info. L, M, C remain as-is; combo = sum of L+M+C for each transect (exclude Other/NA as not appropriate for overall cover density)

# detect which height classes are present in the data, so that we can test even if all combo, or if only L and M present etc.
available_classes <- unique(plant_prop_transect_full_bank$Height_class)

height_sets <- list()
if ("L" %in% available_classes) height_sets$L <- "L"
if ("M" %in% available_classes) height_sets$M <- "M"
if ("C" %in% available_classes) height_sets$C <- "C"

combo_classes <- intersect(c("L","M","C"), available_classes)
if (length(combo_classes) > 0) {
  height_sets$combo <- combo_classes
} else if ("combo" %in% available_classes) {
  # if data already collapsed to "combo"
  height_sets$combo <- "combo"
}

# Function: run GLM per height class and per bank (if available)
run_height_glm <- function(height_name, height_classes, full_df) {
  
  # Filter for the height class(es)
  df_h <- full_df %>%
    filter(Height_class %in% height_classes)
  
  # For combo we need to sum L+M+C for the same transect/site/year
  if (height_name == "combo") {
    df_h <- df_h %>%
      group_by(Site, year, Transect_ID, Bank_group) %>%
      summarise(cover.per.m = sum(cover.per.m), .groups = "drop")
  }
  
  # Add Time_Category
  df_h <- df_h %>%
    mutate(
      Time_Category = case_when(
        year %in% before_years ~ "Before",
        year %in% after_years  ~ "After",
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(Time_Category))
  
  # Determine bank logic
  banks_present <- unique(df_h$Bank_group)
  
  if (any(banks_present %in% c("Left", "Right"))) {
    banks_to_run <- intersect(banks_present, c("Left", "Right"))
  } else {
    banks_to_run <- "Combined"
  }
  
  # Store results
  results <- list()
  diagnostics <- list()
  
  # Run GLM per bank
  for (bank in banks_to_run) {
    
    if (bank != "Combined") {
      df_b <- df_h %>% filter(Bank_group == bank)
    } else {
      df_b <- df_h
    }
    
    # Small constant for Gamma(log) ...not inconceivable that a plant measured at transect A and 'absent' at transect B is actually present but not measured due to lower cover density
    df_b <- df_b %>% mutate(cover.per.m_adj = cover.per.m + 0.001)
    
   if (length(unique(df_b$Site)) < 2 || length(unique(df_b$Time_Category)) < 2) {
  message(paste("Skipping GLM for", height_name, "–", bank,
                "because not enough levels in Site or Time_Category"))
  results[[bank]] <- NA
  diagnostics[[bank]] <- NA
  next
}
    # Fit GLM
    glm_mod <- glm(
      cover.per.m_adj ~ Site * Time_Category,
      data   = df_b,
      family = Gamma(link = "log")
    )
    
    # Store summary
    results[[bank]] <- summary(glm_mod)
    
    # Save to txt file
    out_text <- capture.output(summary(glm_mod))
    export_summary(
      out_text,
      paste0("Total_cover_GLM_output_", bank, "_bank_", height_name, "_height",".txt")
    )
    
    # Diagnostics
    df_b$residuals <- residuals(glm_mod, type = "pearson")
    df_b$fitted    <- fitted(glm_mod)
    
    p_resid <- ggplot(df_b, aes(x = residuals)) +
      geom_histogram(bins = 30, fill = "purple", alpha = 0.7) +
      geom_density(color = "yellow", linewidth = 1) +
      labs(title = paste("Residual Histogram –", height_name, "–", bank),
           x = "Residuals", y = "Frequency") +
      theme_minimal()

    p_fitted <- ggplot(df_b, aes(x = fitted, y = residuals)) +
      geom_point(alpha = 0.5) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "yellow") +
      labs(title = paste("Fitted vs Residuals –", height_name, "–", bank),
           x = "Fitted values", y = "Residuals") +
      theme_minimal()
    
    diagnostics[[bank]] <- list(
      data = df_b,
      resid_plot = p_resid,
      fitted_plot = p_fitted
    )
    
    print(p_resid)
    print(p_fitted)
  }
  
  return(list(results = results, diagnostics = diagnostics))
}

# Run function for all height classes (L, M, C, combo)

glm_outputs <- list()

for (h in names(height_sets)) {
  glm_outputs[[h]] <- run_height_glm(
    height_name    = h,
    height_classes = height_sets[[h]],
    full_df        = plant_prop_transect_full_bank
  )
}

glm_outputs

```

```{r glm interaction only results table, echo=FALSE}
#full results information is generated in above chunk, and model summaries are exported. but for display here, we display only the interactions of all of the bank-height combos that were run (otherwise too cluttered)

extract_interaction <- function(glm_outputs, digits=3) {
  
  results_df <- data.frame(
    Height_Bank = character(),
    Estimate = numeric(),
    Std_Error = numeric(),
    t_value = numeric(),
    p_value = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (height_name in names(glm_outputs)) {
        height_block <- glm_outputs[[height_name]]$results
    
    for (bank_name in names(height_block)) {
      mod_summary <- height_block[[bank_name]]
      if (is.null(mod_summary) || length(mod_summary) == 0) next
      if (is.null(mod_summary$coefficients)) next
      coef_table <- mod_summary$coefficients
      
      interaction_row <- grep("Site.*Time_Category", rownames(coef_table))
      
      if (length(interaction_row) == 1) {
        results_df <- rbind(
          results_df,
          data.frame(
            Height_Bank = paste0(height_name, " – ", bank_name),
            Estimate    = round(coef_table[interaction_row, 1], digits),
            Std_Error   = round(coef_table[interaction_row, 2], digits),
            t_value     = round(coef_table[interaction_row, 3], digits),
            p_value     = round(coef_table[interaction_row, 4], digits),
            stringsAsFactors = FALSE
          )
        )
      }
    }
  }
  
  results_df
}

interaction_results_table <- extract_interaction(glm_outputs)

datatable(
  interaction_results_table,
  escape = FALSE,
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Summary of GLM Site*Time Interactions Only"
  ),
  rownames = FALSE,
  options = list(
    columnDefs = list(
      list(className = "dt-center", targets = "_all")
    ),
    pageLength = 10
  )
)

export_table(as.data.frame(interaction_results_table),
             "GLM_SiteTime_Interaction_Summary.csv")
```


### Species or Feature Cover Density: Generalised Linear Model (GLM)

Now we will apply the same inferential test to individual species (or non-plant features) that you encountered. 

In the next chunk of code you can specify a particular species/feature and a height class (or combine all height classes). A glm will then be run for that species (and height class) so that you can assess the statistical significance of the observed changes in cover density of that species at the restoration site compared with the control or reference site (i.e., the Site*Time interaction). The height class option is, as before, to allow you to assess species as a standard biological concept, or explore the different functions that we have assumed that different heights of plant can play. Note that if you are specifying a non-plant feature, you should set the target_height to "other".

If this seems a bit speculative to you, it probably is. As always, you should only run these tests if you have _a priori_ hypotheses and predictions for the particular species/feature, and should avoid the temptation to dredge through various species looking for any significant changes (which you would surely observe, just due to chance, if you look at a number of species). 

```{r specify species or feature, echo=TRUE}
target_species <- "PHALARU"   # <-- add plant_code value for focal species
target_height <- "L"           # <-- define "C", "M", "L", or "combo" = all height classes, "other" = non-plants
```

```{r glm species density BACI, echo = FALSE}
# apply user specified categories
species_subset <- plant_prop_transect_full_bank %>%
  filter(plant_code == target_species) %>%
  filter(
    case_when(
      tolower(target_height) == "combo" ~ TRUE,
      tolower(target_height) == "other" ~ is.na(Height_class) | Height_class == "Other",
      TRUE ~ Height_class == target_height
    )
  ) %>%
  mutate(Time_Category = case_when(
    year %in% before_years ~ "Before",
    year %in% after_years  ~ "After",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Time_Category))

# ensure banks grouped as specified
banks_present <- unique(species_subset$Bank_group)
if (any(banks_present %in% c("Left", "Right"))) {
  banks_to_run <- intersect(banks_present, c("Left", "Right"))
} else {
  banks_to_run <- "Combined"
}

# run GLM per bank if present
results_list <- list()
diagnostics_list <- list()

for (bank in banks_to_run) {
  
  if (bank != "Combined") {
    species_bank <- species_subset %>% filter(Bank_group == bank)
  } else {
    species_bank <- species_subset
  }
  
  # Choose GLM family depending on the data
  # cover.per.m has zeros and can be >1, so Gamma with log link can work.
  # or a quasipoisson for count-like continuous response, or the more general Tweedie (can include zeros)
  # Here, we add a small constant so we can use Gamma (to avoid zeros)
  species_bank <- species_bank %>%
    mutate(cover.per.m_adj = cover.per.m + 0.001) #not inconceivable that a plant measured at transect A and 'absent' at transect B is actually present but not measured due to lower cover density
  
  # Fit GLM
  glm_mod <- glm(
    cover.per.m_adj ~ Site * Time_Category,
    data = species_bank,
    family = Gamma(link = "log")  # an alternative is Tweedie (more generalised distribution, handles zero)
  )
  
  results_list[[bank]] <- summary(glm_mod)
  
  out_text <- capture.output(summary(glm_mod))
  export_summary(
    out_text,
    paste0("GLM_CoverDensity_", target_species, "_Height_", target_height, "_Bank_", bank, ".txt")
  )
}

# Diagnostics 
  species_bank$residuals <- residuals(glm_mod, type = "pearson")
  species_bank$fitted <- fitted(glm_mod)
  
  p_resid <- ggplot(species_bank, aes(x = residuals)) +
    geom_histogram(bins = 30, fill = "purple", alpha = 0.7) +
    geom_density(color = "yellow", linewidth = 1) +
    labs(title = paste("Residuals Histogram – Bank", bank),
         x = "Pearson Residuals", y = "Frequency") +
    theme_minimal()
  
  p_fitted <- ggplot(species_bank, aes(x = fitted, y = residuals)) +
    geom_point(alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "yellow") +
    labs(title = paste("Fitted vs Residuals – Bank", bank),
         x = "Fitted Values", y = "Residuals") +
    theme_minimal()
  
  diagnostics_list[[bank]] <- list(
    data = species_bank,
    residual_plot = p_resid,
    fitted_vs_resid_plot = p_fitted
  )

  print(p_resid)
  print(p_fitted)


results_list        
diagnostics_list   
```

So that is it for cover density for now. As mentioned previously, we hope to return to this code to refine it. Until then, we hope it is useful. For any future refinements we will ensure we keep using the same metrics calculated in the same way (so that it is comparable to anything generated now) but we hope to add to the metrics and provide more options, as well as improving the inferential testing we offer.

Next we will move onto some characteristics of the vegetation: wetland (hydrophyte) status, invasive status, and overhanging extents of the vegetation.

### R and Package Version Information

We provide the following information for the sake of reproducability. Some future updates may involve changes that impact the ability to run this script, so it is useful to know which versions we used so that you can continue to use it.

```{r session info, echo = TRUE}
sessionInfo()
```